---
title: "13: High Frequency Data"
author: "Hydrologic Data Analysis | Cathy Chamberlin"
date: "Fall 2019"
output: html_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Lesson Objectives
1. Determine stormflow and baseflow from high frequency flow data
2. Use a hysteresis plot to understand watershed dynamics
3. Communicate findings with peers through oral, visual, and written modes

## Opening Discussion

High frequency data is usually defined as frequencies significantly lower than daily (e.g. 5-minute, 15-minute, 1 hr etc). What types of hydrological and biological processes happen on this timescale that we might want to investigate?

>


## Session Set Up

```{r, message = FALSE}
getwd()

#install.packages("EcoHydRology")

packages <- c(
  "dataRetrieval", "tidyverse", "cowplot", "EcoHydRology", "xts", "dygraphs"
  )
invisible(lapply(packages, library, character.only = TRUE)) 

theme_set(theme_classic(base_size = 12))
```

## Pull Data

```{r download data}

dischargedatTFC2018 <- readNWISuv(
  site = "02097280", #Third Fork Creek at Woodcroft Parkway Near Blands
  parameterCd = "00060",
  startDate = "2018-01-01", 
  endDate = "2019-01-01"
  ) %>% 
  renameNWISColumns()


dischargedatEC2018 <- readNWISuv(
  site = "0208675010", #Ellerbe Creek at Club Boulevard at Durham, NC
  parameterCd = "00060", 
  startDate = "2018-01-01", 
  endDate = "2019-01-01"
  ) %>% 
  renameNWISColumns()


plot_grid(
  ggplot(dischargedatTFC2018, aes(x = dateTime, y = Flow_Inst)) +
    geom_line() + 
    scale_y_log10() +
    ggtitle("Third Fork Creek"),
  ggplot(dischargedatEC2018, aes(x = dateTime, y = Flow_Inst)) + 
    geom_line() + 
    scale_y_log10() +
    ggtitle("Ellerbe Creek"),
  ncol = 1
)
          

```

The shape and size of river hydrographs following rain events can tell us how water moves through a watershed. Once rain falls, how does water get to a stream? What would the streamflow do?

> 


_"Hydrograph separation is one of the most desperate analysis techniques in use in hydrology" - Hewlett & Hibbert 1967_


```{r baseflowseparation}
TFCbaseflow <- BaseflowSeparation(
  dischargedatTFC2018$Flow_Inst, 
  filter_parameter = 0.925, 
  passes = 3
  )

# the filter_parameter and passes functions are set to the default

TFC2018 <- cbind(dischargedatTFC2018, TFCbaseflow)

ggplot(TFC2018, aes(x = dateTime, y = Flow_Inst)) + 
  geom_line() +
   scale_y_log10() +
  geom_line(mapping = aes(x = dateTime, y = bt), color = "darkorange4") +
  geom_line(mapping = aes(x = dateTime, y = qft), color = "steelblue4")

# making it log scale doesn't work so well because there are a bunch of 0s in the data

dygraph(
  cbind(
    Flow = with(TFC2018, xts(Flow_Inst, order.by = dateTime)), 
    Baseflow = with(TFC2018, xts(bt, order.by = dateTime)), 
    Quickflow = with(TFC2018, xts(qft, order.by = dateTime))
    )
  ) %>% 
  dyRangeSelector()

Export <- TFC2018 %>%
  mutate(timestep = c(diff(as.numeric(dateTime)), NA_real_),
# use diff function to find the difference in time between each data point
# making dateTime numeric gives us time in seconds. We want it in seconds because our discharge is in cfs
# NA_real_ creates an NA value in order for diff to produce an equal number of rows to our dataset (in the summarize function below, we remove the NA)
         baseflowexport = bt * timestep,
         quickflowexport = qft * timestep) %>%
# the above equations give us the volume of water between each data point
  summarize(BaseflowExport_cf = sum(baseflowexport, na.rm = T),
            QuickflowExport_cf = sum(quickflowexport, na.rm = T),
            TotalExport_cf = BaseflowExport_cf + QuickflowExport_cf)
  
```

What percentage of water is exported from this watershed as baseflow? What percentage as quickflow?

> 51.9% is baseflow, 48.1% is quick flow

Repeat this analysis for Ellerbe Creek. How do the percentages compare?

> 43.4% is baseflow, 56.6% is quick flow

```{r}
ECbaseflow <- BaseflowSeparation(
  dischargedatEC2018$Flow_Inst, 
  filter_parameter = 0.925, 
  passes = 3
  )

EC2018 <- cbind(dischargedatEC2018, ECbaseflow)

ExportEC <- EC2018 %>%
  mutate(timestep = c(diff(as.numeric(dateTime)), NA_real_),
         baseflowexport = bt * timestep,
         quickflowexport = qft * timestep) %>%
  summarize(BaseflowExport_cf = sum(baseflowexport, na.rm = T),
            QuickflowExport_cf = sum(quickflowexport, na.rm = T),
            TotalExport_cf = BaseflowExport_cf + QuickflowExport_cf)
```


What does this mean physically about the watershed?

> The Ellerbe Creek watershed is likely more urbanized than the Third Fork Creek watershed.

## Hydrologic Flashiness

A "flashy" stream increases in flow faster following a precipitation event. 

Refering to our previous work with baseflow separation, would you expect flashiness to increase or decrease with urbanization?

> Flashiness would increase with urbanization.

What do you think "urban stream syndrom" refers to?

> Urban streams have higher quick flow due to more impervious surfaces relative to natural conditions.

The Richards-Baker Index is an index of how flashy a stream is. It is calculated as $$RBI = (\sum_{j = 1}^n |Q_j - Q_{j-1}| / \sum_{j = 1}^n Q_j) / Catchment Area$$

  - j = timestep

```{r Flashiness}
TFCsitedata <- readNWISsite(site = "02097280")
TFC.catchment.size <- TFCsitedata$drain_area_va #square miles

RBI_TFC2018 <- (
  sum(abs(diff(dischargedatTFC2018$Flow_Inst))) / 
    sum(dischargedatTFC2018$Flow_Inst[-1])
  ) / TFC.catchment.size

ECsitedata <- readNWISsite(site = "0208675010")
EC.catchment.size <- ECsitedata$drain_area_va #square miles

RBI_EC2018 <- (
  sum(abs(diff(dischargedatEC2018$Flow_Inst))) / 
    sum(dischargedatEC2018$Flow_Inst[-1])
  ) / EC.catchment.size

```

Which stream is flashier?

> Ellerbe Creek is flashier because the RBI (0.012) is higher than the RBI for Third Fork Creek (0.003)

## Chemical Flashiness

Chemical concentrations in rivers can be either highly variable with discharge, or "chemostatic", which means the concentration barely changes with discharge. 

```{r}
CCdat <- readNWISuv(
  site = "02249500", # CRANE CREEK AT MELBOURNE, FL
  parameterCd = c("00060", "99133"), # Discharge in cfs & Nitrate in mg/l NO3-N
  startDate = "2015-01-1",
  endDate = "2016-01-01"
  ) %>%
  renameNWISColumns() %>%
  rename(Nitrate_mgl = 6)

ggplot(CCdat,
       aes(x = Flow_Inst, y = Nitrate_mgl)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()


Potdat <- readNWISuv(
  site = "01646500", # POTOMAC RIVER NEAR WASH, DC LITTLE FALLS PUMP STA
  parameterCd = c("00060", "99133"), # Discharge in cfs & Nitrate in mg/l NO3-N
  startDate = "2016-01-1",
  endDate = "2017-01-01"
  ) %>% 
  renameNWISColumns() %>%
  rename(Nitrate_mgl = 6)

ggplot(Potdat,
       aes(x = Flow_Inst, y = Nitrate_mgl)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

Which of these rivers is more chemostatic?

>

## Hysteresis Loops

Hysteresis occurs when the concentration varies with discharge in a different relationship on the rising and falling limbs of a storm. This can be caused by _flushing_ behavior or _dilution_ behavior. 

```{r}
dygraph(
  cbind(
    Flow = xts(Potdat$Flow_Inst, order.by = Potdat$dateTime), 
    Nitrate = xts(Potdat$Nitrate_mgl, order.by = Potdat$dateTime)
    )
  ) %>% 
  dySeries("Nitrate", axis = "y2") %>%
  dyRangeSelector()
```

Look at the storm that occured around February 1st. Do you expect this storm to be a flushing or a diluting storm?

>

```{r}
PotomacStorm <- Potdat %>%
  filter(dateTime > "2016-01-31" & dateTime < "2016-02-14") 

ggplot(PotomacStorm, aes(x = Flow_Inst, y = Nitrate_mgl, color = dateTime)) +
  geom_point() 


```

Did this storm exhibit clockwise or counterclockwise hysteresis?

>

Which direction do hysteresis loops go for diluting storms?

>

Which direction do hysteresis loops go for flushing storms?

>

## Closing Discussion

What hydrological processes did we investigate today? 

>

What can high-frequency data be used for that daily data may not be able to capture?

> 
